Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package

### 2021-05-06 010-background.Rmd
Drafting comparison table in background v2

```{r tab:compare-table}
knitr::kable(
           head(mtcars[, 1:8], 10),  booktabs = TRUE, format = "latex",
                 linesep = c("", "", "\\addlinespace"),
           caption = 'A table of the first 10 rows of the mtcars data.'
  )
```

<!-- 
| Software | Designed for | Scales to Large P | Hetereogenous Margins | Flexible Dependence |
|----------|--------------|-------------------|-----------------------|---------------------|
|`copula`|              |                   |                       |                     |
|`genord`|              |                   |                       |                     |
|`mvnfast`|              |                   |                       |                     |

your comment -->

>>>>

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, as well as make this work's contribution more evident, we make some qualitative comparisons to other `R` multivariate simulation packages.  The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. Recently, the high-performance package `mvnfast` has been released that provides scaleable multivariate normal generation, but this doesn't completely satistify our goals of flexible margins and dependency modeling. For dependent discrete data, small-dimesional packages exist, such as `GenOrd`. The full-featured package `copula` provides many different copulas and a fairly general random generator `rCopula`. For low-dimensional modeling situtations, copula is suitable, yet the entent of exact matching of margins automatically is unclear. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for low-dimenional settings. In sum, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table (\#tab:compare-table) summarizes the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
\label{tab:compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

### 2021-05-06 040-monte-carlo-evaluation.Rmd
Drafting comparison table in background v2

\begin{table}[]
\begin{tabular}{@{}lcr@{}}
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}

@@

\setstretch{1.5}
i. Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
1. Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.
2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.
3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.
\setstretch{2.0}

@@

```{r ch040-moderateDtab}
dat_long <- benchmark_dependences %>%
  select(-total_time, -n_sim, -needed_near_pd) %>%
  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
  mutate(dim = factor(dim),
         Step = factor(Step,
                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
                       labels = c("Compute Correlation",
                                  "Adjust Correlation",
                                  "Check Admissibility",
                                  "Simulate Data"))) %>%
    rename(Correlation = corr_type, Dimensions = dim)

@@

tabTimes %>%
    kable(booktabs = TRUE,
          linesep = c("", "", "\\addlinespace"),
          col.names = c("Dimension",
                        "Correlation type",
                        "Total Time (Seconds)"),
          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

>>>>

\begin{table}[]
\begin{tabular}{@{}lcr@{}}
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}

@@

\setstretch{1.5}
1. a) Estimate the correlation matrix from the "data" in the *Compute Correlation* step.
1. b) Map the correlations to initialize the algorithm (Pearson to Pearson, Spearman to Pearson, or Kendall to Pearson) in the *Adjust Correlation* step.
2. Check whether the mapping produces a valid correlation matrix and, if not, find the nearest PD correlation matrix in the *Check Admissibility* step.
3. Simulate $1,000$ vectors via the Inverse Transform method in the *Simulate Data* step.
\setstretch{2.0}

@@

```{r tab:ch040-moderateDtab}
dat_long <- benchmark_dependences %>%
  select(-total_time, -n_sim, -needed_near_pd) %>%
  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
  mutate(dim = factor(dim),
         Step = factor(Step,
                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
                       labels = c("Compute Correlation",
                                  "Adjust Correlation",
                                  "Check Admissibility",
                                  "Simulate Data"))) %>%
    rename(Correlation = corr_type, Dimensions = dim)

@@

tabTimes %>%
    kable(booktabs = TRUE,
          linesep = c("", "", "\\addlinespace"), format="latex",
          col.names = c("Dimension",
                        "Correlation type",
                        "Total Time (Seconds)"),
          caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
```

### 2021-05-06 article_bigsimr.tex
Drafting comparison table in background v2

\caption{\label{tab:tab:compare-table}A table of the first 10 rows of the mtcars data.}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
  & mpg & cyl & disp & hp & drat & wt & qsec & vs\\
\midrule
Mazda RX4 & 21.0 & 6 & 160.0 & 110 & 3.90 & 2.620 & 16.46 & 0\\
Mazda RX4 Wag & 21.0 & 6 & 160.0 & 110 & 3.90 & 2.875 & 17.02 & 0\\
Datsun 710 & 22.8 & 4 & 108.0 & 93 & 3.85 & 2.320 & 18.61 & 1\\
\addlinespace
Hornet 4 Drive & 21.4 & 6 & 258.0 & 110 & 3.08 & 3.215 & 19.44 & 1\\
Hornet Sportabout & 18.7 & 8 & 360.0 & 175 & 3.15 & 3.440 & 17.02 & 0\\
Valiant & 18.1 & 6 & 225.0 & 105 & 2.76 & 3.460 & 20.22 & 1\\
\addlinespace
Duster 360 & 14.3 & 8 & 360.0 & 245 & 3.21 & 3.570 & 15.84 & 0\\
Merc 240D & 24.4 & 4 & 146.7 & 62 & 3.69 & 3.190 & 20.00 & 1\\
Merc 230 & 22.8 & 4 & 140.8 & 95 & 3.92 & 3.150 & 22.90 & 1\\
\addlinespace
Merc 280 & 19.2 & 6 & 167.6 & 123 & 3.92 & 3.440 & 18.30 & 1\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  Estimate the correlation matrix from the ``data'' in the \emph{Compute
  Correlation} step.
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Map the correlations to initialize the algorithm (Pearson to Pearson,
  Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust
  Correlation} step.
\item
  Check whether the mapping produces a valid correlation matrix and, if
  not, find the nearest PD correlation matrix in the \emph{Check
  Admissibility} step.
\item
  Simulate \(1,000\) vectors via the Inverse Transform method in the
  \emph{Simulate Data} step. \setstretch{2.0}
\end{enumerate}

@@

Table: Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.

|Dimension |Correlation type | Total Time (Seconds)|
|:---------|:----------------|--------------------:|
|100       |Pearson          |                0.081|
|100       |Spearman         |                0.027|
|100       |Kendall          |                0.075|
|250       |Pearson          |                0.378|
|250       |Spearman         |                0.062|
|250       |Kendall          |                0.375|
|500       |Pearson          |                1.425|
|500       |Spearman         |                0.135|
|500       |Kendall          |                1.495|

@@

\hypertarget{examples}{%
\section{RNA-seq data applications}\label{examples}}

>>>>

To place \texttt{Bigsimr}'s implementation and algorithms in context
with existing approaches, as well as make this work's contribution more
evident, we make some qualitative comparisons to other \texttt{R}
multivariate simulation packages. The \texttt{MASS} packages
\texttt{mvnorm()} has long been available to produce random vectors from
multivariate normal (and \(t\)) distributions. This familiar procedure
cannot simulate from arbitrary margins and dependency measures, as well
as fails to readily scale to high dimensions. Recently, the
high-performance package \texttt{mvnfast} has been released that
provides scaleable multivariate normal generation, but this doesn't
completely satistify our goals of flexible margins and dependency
modeling. For dependent discrete data, small-dimesional packages exist,
such as \texttt{GenOrd}. The full-featured package \texttt{copula}
provides many different copulas and a fairly general random generator
\texttt{rCopula}. For low-dimensional modeling situtations, copula is
suitable, yet the entent of exact matching of margins automatically is
unclear. High-dimesional random vectors via \texttt{copula} takes an
impractical amount of computing time \citep{Li2019gpu}. The
\texttt{nortaRA} R package matches Pearson coefficients near exactly and
is reasonable for low-dimenional settings. In sum, we know of no other
\texttt{R} package that can produce HD random vectors with flexible
margins and dependence modeling. Table (\#tab:compare-table) summarizes
the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
\label{tab:compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \tightlist
  \item
    Estimate the correlation matrix from the ``data'' in the
    \emph{Compute Correlation} step.
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \setcounter{enumii}{1}
  \tightlist
  \item
    Map the correlations to initialize the algorithm (Pearson to
    Pearson, Spearman to Pearson, or Kendall to Pearson) in the
    \emph{Adjust Correlation} step.
  \end{enumerate}
\item
  Check whether the mapping produces a valid correlation matrix and, if
  not, find the nearest PD correlation matrix in the \emph{Check
  Admissibility} step.
\item
  Simulate \(1,000\) vectors via the Inverse Transform method in the
  \emph{Simulate Data} step. \setstretch{2.0}
\end{enumerate}

@@

\caption{\label{tab:tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
\centering
\begin{tabular}[t]{llr}
\toprule
Dimension & Correlation type & Total Time (Seconds)\\
\midrule
100 & Pearson & 0.081\\
100 & Spearman & 0.027\\
100 & Kendall & 0.075\\
\addlinespace
250 & Pearson & 0.378\\
250 & Spearman & 0.062\\
250 & Kendall & 0.375\\
\addlinespace
500 & Pearson & 1.425\\
500 & Spearman & 0.135\\
500 & Kendall & 1.495\\
\bottomrule
\end{tabular}
\end{table}

@@

\hypertarget{examples}{%
\section{RNA-seq data application}\label{examples}}
